---
title: "R Practice"
author: "NDP"
date: "6/2/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Intro to markdown

Markdown allows you to do things like **bold**, *italic*, `code font`
and more!

Make a list with asterisks like:

-   \* bold with double-asterisk

-   \* italics with underscores

-   \* code-type fond with backticks

-   You can also use hyphens instead of asterisks.

For a numbere list:

1.  1\. bold.....

2.  1\. italic....

3.  1\. code.....

Hyperlinks are like [Hyperlinked text](https://mylink.org)


Load gapminder data in along with plotting package.

```{r load_data}
#This line will not run
library(gapminder) #this chunk of a line will not run
library(ggplot2)
gap <- gapminder
```

Try and fail to plot the data.

```{r make_plot}
plot(lifeExp ~ year, data=gap)
```

```{r}
temp <- c(1,2,3) 
```

```{r make_cats}
cats <- data.frame(coat = c("calico","black","tabby"),
                   weight=c(2.1, 5.0, 3.2),
                   likes_string = c(1, 0, 1))
write.csv(x = cats, file = "data/feline-data.csv", row.names=FALSE)
```



```{r}
library(rstudioapi)
setwd(dirname(getActiveDocumentContext()$path))
```

```{r read_cats}
cats2 <- read.csv('data/feline-data.csv', stringsAsFactors = Ts)
```

Creating vectors and coercing types

```{r}
my_vector <- vector(mode='character',length=3)
combine_vector <- c(1,2,"bananas")
```

change names

```{r}
names(cats)[3] <- "likes_hunting" 
```

Add columns and rows to a df

```{r}
cats <- read.csv('data/feline-data.csv')
age <- c(2,3,5)
cats <- cbind(cats,age)
newRow <- list("tortoiseshell", 3.3, TRUE, 9)
cats <- rbind(cats,newRow)
```

CHALLENGE:
Create a vector called human_age by multiplying cats$age by 7.
Convert human_age to a factor.
Convert human_age back to a numeric vector using the as.numeric() function. Now divide it by 7 to get the original ages back. Explain what happened.

```{r}
human_age <- cats$age * 7 #create human_age
human_age <- as.factor(human_age) #convert to factor
human_age <- as.numeric(as.character(human_age)) #convert to numeric w/as.numeric()
human_age/7 #divide by 7
```

Remove rows and columns

```{r}
cats <- read.csv('data/feline-data.csv')
cats$coat <- as.factor(cats$coat)
cats <- rbind(cats,c("white",3.4,0,1))
#remove a row
cats <- cats[-5,]
#cats <- na.omit(cats)
cats[,-4]

drop <- names(cats) %in% c("weight")
cats[,!drop]
```
Insert a code chunk using a keyboard shortcut (`CTRL-ALT-I` _or_ `CMD-OPT-I`)

```{r}
library(readxl)
#download at https://github.com/ndporter/dspg-r-2022/raw/gh-pages/data/gapminder_data.xlsx
gapminder <- read_excel("data/gapminder_data.xlsx")
#you can use download.file

```

Inspecting a dataframe

```{r}
summary(gapminder)
summary(gapminder$lifeExp)

gapminder$country <- factor(gapminder$country)
gapminder$continent <- factor(gapminder$continent)
summary(gapminder)

length(gapminder) #ncol(gapminder)
nrow(gapminder)
colnames(gapminder)

head(gapminder)
tail(gapminder)
gapminder[sample(nrow(gapminder), 10), ]
gapminder[1:10,2:5]
```
## Subsetting Data

```{r}
x <- c(5.4,6.2,7.1,4.8,7.5)
names(x) <- c('a','b','c','d','e')

x[c(1,3)] #discrete items
x[1:3] #slice
x[c(1,1,3)]
x[6]#NA - beyond the end
x[-2]
x[c(-1,-5)]
x[-c(1,5)]
x[-(1:3)]
x <- x[-4]
#by name
x[c("a","c")]

x <- c(5.4,6.2,7.1,4.8,7.5)
names(x) <- c('a','b','c','d','e')
x[x>7 & x<7.5] #and
x[x>7 | x<7.5] #or

x2 <- 1:3
names(x2) <- c('a','a','a')
x2['a'] #duplicate names are possible but not advisable

x
x[! names(x) %in% c("a","c")]
```

Select southeast Asian countries from gapminder

```{r}
seAsia <- c("Myanmar", "Thailand", "Cambodia", "Vietnam", "Laos")
gapminder[which(gapminder$country %in% seAsia),]
seAsia_gap <- gapminder[gapminder$country %in% seAsia,]
```

Extract observations from the 1960s

```{r}
gap_60s <- gapminder[gapminder$year>1959 & gapminder$year<1970,]
```

**PICTURES!!!!!!!!!**

```{r}
library("ggplot2")
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + geom_point()
```
```{r}
#this creates axis, not figure
ggplot(data=gapminder, mapping = aes(x=gdpPercap, y=lifeExp))
```
```{r}
ggplot(data=gapminder, mapping = aes(x=year, y=lifeExp)) +
  geom_point()

```
Lines!

```{r}
ggplot(data=gapminder, mapping = aes(x=year, y=lifeExp, color=continent)) +
  geom_line()
```
LInes that make a little sense

```{r}

ggplot(data=gapminder, mapping = aes(x=year, y=lifeExp, by=country)) +
  geom_line(mapping = aes(color=continent)) + geom_point(color="yellow")
```

```{r}
logged_points <- ggplot(gapminder, mapping= aes(x=gdpPercap, y=lifeExp)) +
  geom_point()+ scale_x_log10()
```

```{r}
logged_points + geom_smooth(method='loess')
```

```{r}
americas <- gapminder[gapminder$continent=="Americas",]
ggplot(data=americas, mapping = aes(x=year, y=lifeExp)) +
  geom_line() +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle=45))

ggsave(filename='results/lifeExp_americas.png')
```
Save Americas!

```{r}
write.csv(file = 'clean_data/americas_gap.csv', x=americas, row.names=F)
```

dplyr!

```{r}
#tried to get mean for africa
#mean(gapminder[factor(gapminder$continent=="Africa"),"gdpPercap"])
```

```{r}
library("dplyr")
year_country_gdp <- select(gapminder, year, country, gdpPercap)
str(year_country_gdp)
smaller_gap <- gapminder %>% select(-continent) %>% rename(gdp_per_capita = gdpPercap)
```
```{r}
y_cntr_gdp_eur <- gapminder %>%
  filter(continent=="Europe") %>%
  select(year, country, gdpPercap)
```

Combine with grouping

```{r}
gapminder %>% group_by(continent, year) %>% 
  summarize(mean_gdpPercap = mean(gdpPercap),
            sd_gdpPercap = sd(gdpPercap),
            mean_raw_gdp = mean(gdpPercap*pop),
            tot_raw_gdp = sum(gdpPercap*pop))
```
Mutate!

```{r}
gapminder %>% mutate(gdp_billion = gdpPercap*pop/10^9) %>%
  group_by(continent, year) %>%
  summarize(mean_gdp_bill = mean(gdp_billion))
```
  
Comabine with ggplot

```{r}
gapminder %>%
  filter(continent=="Americas") %>%
  ggplot(mapping = aes(x=year, y=lifeExp)) +
  geom_line() + 
  facet_wrap( ~ country)
```

